Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-18 jss/EpiCompare-paper/EpiCompare-paper.Rmd
thursday chat edits

The first step of most data analysis is cleaning the data \ben{so it can \sout{to}} be explored. There are multiple ways to collect epidemiological data. Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections focus on aggregate counts of individuals in each epidemic state.  In fact, usually only the number of new infections at each time step (e.g. weekly case counts) is observed.  Compartment totals \ben{(amounts of individuals in each state)} are then imputed from those case counts along with other information about the disease and the population of interest.  \ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows examination of overall trends of an epidemic more easily}.

@@

\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc) to an aggregate view of how many individuals where in a state at a given time. There are often situations where grouping agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for subpopulations in their models\footnote{\ben{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}. This function allows the user to easily group agents into new subpopulations by using the \pkg{tidyverse} paradigm to first `group_by()`^[The `group_by` function is from the \pkg{dplyr} package.] the subpopulation defining features and then aggregate (using `agents_to_aggregate`) conditional on these subpopulations.  \shannon{I think these last two sentences are repetitive}

\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is also flexible to deal with a wide range of information about each individual. It can, theoretically, account for infinitely many states, allowing the practitioner to aggregate information relative to the standard "Suspectible", "Infectious", and "Recovered" states (CITE) or also add states like "Exposed", "iMmune", "Hospitalized" and more. Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates (that change the overall population size). \shannon{Don't think this is necessary \sout{This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.}} The function `agents_to_aggregate()` is a powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.

@@

\ben{[Ben says: this is actually an older version:]} With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets. An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but \pkg{EpiCompare} also includes tools to rapidly explore your data if your has at least three epidemic states (although these tools work best with exactly three epidemic states). Our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of the epidemic, and combines the ideas behind `agents_to_aggregate` for the SIR case, with `geom_path` and `coord_tern` to visualize any number of groups epidemic trajectory in 3d simplex space using \pkg{ggplot2} and \pkg{ggtern} \citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).

@@

**Model Fitting and Simulations**
\ben{[Ben says: see footnote 3 about the introduction to this paragraph.]}

Although this package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.\sout{ with Bernoulli or Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next}^[The original sentence seemed clearer to me (it may even be a run-on sentence now). I'm wondering if you didn't like the original sentence relative to the "truth-ful-ness" of the statement? Here's the original sentence slightly edited: "Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents".].  These functions can be naturally combined with `agents_to_aggregate()` to proceed in the pipeline.  The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.  This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

>>>>

The first step of most data analysis is cleaning the data \ben{so it can \sout{to}} be explored. There are multiple ways to collect epidemiological data. Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections focus on aggregate counts of individuals in each epidemic state.  In fact, usually only the number of new infections at each time step (e.g. weekly case counts) is observed.  Compartment totals \ben{(amounts of individuals in each state)} are then imputed from those case counts along with other information about the disease and the population of interest.  \ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows for examination of overall trends of an epidemic more easily}.

@@

\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. There are often situations where grouping agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining \pkg{dplyr}::`group_by` and `agent_to_aggregate` to provide aggregation at a group level.

\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is flexible and can deal with a wide range of information about each individual. It can, theoretically, account for infinitely many states.  This functionality allows the practitioner to aggregate information relative to the standard states (e.g. "Susceptible", "Infectious", and "Recovered") as well as add states (e.g. "Exposed", "iMmune", "Hospitalized") (CITE). Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates. Overall, the function `agents_to_aggregate()` is a powerful tool for pre-processing data.

@@

\ben{[Ben says: this is actually an older version:]} With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets. An expert coder has many ways to successfully explore the data in an aggregate lens using `agents_to_aggregate`. \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of an epidemic by combining the ideas behind `agents_to_aggregate` for three-state models to examine subpopulation trajectories in a 3d simplex space \citep{Wickham2016, Hamilton2018}. \shannon{come back?} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}). \shannon{make sure SIR is defined before}

@@

**Model Fitting and Simulations**

Although this package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.  These functions produce individual level information and can be naturally combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
